---
title: "Getting Started with cosTapered"
author: "Andrew O. Finley, Lu Zhang, Arne Nothdurft, and Sudipto Banerjee"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with cosTapered}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Overview

`cosTapered` fits Bayesian change-of-support (COS) models for spatial data
whose responses, covariates, and prediction targets are measured on different
supports. In the COS setting considered here, covariates are represented on a
fine raster grid, responses are observed on polygons, and predictions can be
made for either fine raster cells or other polygon supports.

Development of the package was motivated by data structures common in natural
resource and forest inventory applications, where field responses are measured
on plots, remotely sensed covariates are available as raster layers, and
management decisions often require predictions for stands or other irregular
areal units. The model follows the tapered COS formulation used by Zhang et
al. (2024) for incompatible spatial forest damage data.

The main workflow is:

1. Supply the raster covariate stack used to form the design matrix.
2. Prepare observed-support weights with `cos_prepare()`.
3. Specify priors with `cos_default_priors()`.
4. Fit covariance parameters with `cos_fit()`.
5. Recover observed-support latent quantities with `cos_recover_B()`.
6. Predict fine-support or areal-support quantities with
   `cos_predict_fine()` and `cos_predict_areal()`.

The workflow keeps data preparation choices explicit. The raster stack defines
the design matrix, including any intercept or transformed covariates; observed
supports are supplied as polygons; the taper distance `gamma` is chosen before
fitting; and areal predictions are returned as support averages, leaving any
conversion to totals to the analysis context.

## Example data

The package includes a small simulated example with a canopy-height raster,
observed plot polygons, areal prediction polygons, a forest boundary, and
known simulation truth.

```{r packages}
library(sf)
library(raster)
library(ggplot2)
library(cosTapered)
```

```{r read-data}
chm <- raster(system.file("extdata", "example_chm.tif", package = "cosTapered"))
names(chm) <- "chm"

B_sf <- st_read(system.file("extdata", "example_B.gpkg",
                            package = "cosTapered"), quiet = TRUE)
U_sf <- st_read(system.file("extdata", "example_U.gpkg",
                            package = "cosTapered"), quiet = TRUE)
forest_sf <- st_read(system.file("extdata", "example_forest.gpkg",
                                 package = "cosTapered"), quiet = TRUE)
truth <- readRDS(system.file("extdata", "example_truth.rds",
                             package = "cosTapered"))

truth[c("beta_true", "tau_B_sq_true", "sigma_sq_true",
        "eff_range_true", "phi_true", "gamma", "n_B", "n_U")]
```

## Supply the raster design

The model design matrix is taken from the layers in the raster stack. Here the
stack contains an intercept layer and the canopy-height layer.

```{r build-raster-design}
intercept <- chm
intercept[] <- ifelse(is.na(getValues(chm)), NA_real_, 1)
names(intercept) <- "intercept"

X_rast <- stack(intercept, chm)
names(X_rast) <- c("intercept", "chm")
```

```{r input-plot}
chm_df <- as.data.frame(chm, xy = TRUE, na.rm = TRUE)
names(chm_df) <- c("x", "y", "chm")

ggplot() +
  geom_tile(data = chm_df, aes(x = x, y = y, fill = chm),
            width = xres(chm), height = yres(chm)) +
  geom_sf(data = U_sf, fill = NA, color = "white", linewidth = 0.25) +
  geom_sf(data = forest_sf, fill = NA, color = "black", linewidth = 0.55) +
  geom_sf(data = B_sf, fill = NA, color = "#c1121f", linewidth = 0.45) +
  scale_fill_viridis_c(option = "cividis") +
  coord_sf(expand = FALSE) +
  labs(title = "Raster covariate, observed plots, and prediction polygons",
       fill = "CHM", x = NULL, y = NULL) +
  theme_bw()
```

## Prepare observed supports

`cos_prepare()` computes area weights linking observed polygons to raster
cells, observed-support covariates, the observed-support nugget matrix, and
the distance/taper information needed for spatial fitting.

This is the first model-building step after the raster design and observed
polygons have been assembled.

```{r prepare, eval = FALSE}
prep <- cos_prepare(
  X_rast = X_rast,
  B_sf = B_sf,
  response_col = "y_B",
  gamma = truth$gamma,
  taper_code = truth$taper_code,
  n_threads = 1,
  verbose = FALSE
)

summary(prep)
```

## Specify priors

Priors are specified on the covariate scale supplied by the user. The package
uses `tau_B_sq`, the average observed-support nugget variance, because this is
usually easier to think about than the fine-support nugget variance `tau_sq`.

```{r priors, eval = FALSE}
priors <- cos_default_priors(
  prep = prep,
  beta_mu = c(intercept = 40, chm = 1),
  beta_sd = c(intercept = 20, chm = 2),
  tau_B_shape = 2,
  tau_B_scale = truth$tau_B_sq_true,
  sigma_shape = 2,
  sigma_scale = truth$sigma_sq_true,
  phi_lower = 3 / 600,
  phi_upper = 3 / 80
)
```

The exponential effective range is approximately `3 / phi`. Thus
`phi_lower = 3 / 600` and `phi_upper = 3 / 80` allow effective ranges from
about 80 to 600 map units.

## Fit the model

`cos_fit()` samples the collapsed posterior for covariance parameters. Use
longer chains and multiple starting values for real analyses.

```{r fit, eval = FALSE}
fit <- cos_fit(
  prep = prep,
  priors = priors,
  n_chains = 3,
  starting = data.frame(
    tau_B_sq = truth$tau_B_sq_true * c(0.5, 1, 2),
    sigma_sq = truth$sigma_sq_true * c(0.5, 1, 2),
    phi = truth$phi_true * c(0.75, 1, 1.25)
  ),
  tuning = c(log_tau_B_sq = 0.25, log_sigma_sq = 0.25, z_phi = 0.25),
  n_batch = 100,
  batch_length = 25,
  seed = 1,
  verbose = TRUE
)

summary(fit, burn_in = 50)
plot(fit, burn_in = 50)
```

Trace plots should move across stable regions after burn-in. If `phi` piles up
against a prior bound, revisit the effective range bounds or the taper
distance `gamma`.

## Recover observed-support quantities

`cos_recover_B()` samples regression coefficients, observed-support spatial
effects, and observed-support latent means conditional on retained covariance
parameter draws.

```{r recover, eval = FALSE}
rec_B <- cos_recover_B(
  fit = fit,
  burn_in = 50,
  n_samples = 100,
  seed = 1,
  verbose = FALSE
)

summary(rec_B, include_eta_B = TRUE)
```

## Predict fine or areal supports

Fine-support prediction returns the latent raster-scale response by default.

```{r predict-fine, eval = FALSE}
pred_fine <- cos_predict_fine(
  fit = fit,
  rec_B = rec_B,
  method = "mean",
  verbose = FALSE
)

summary(pred_fine)
```

Areal prediction first builds prediction-support blocks and then predicts
support averages.

```{r predict-areal, eval = FALSE}
U_blocks <- cos_make_areal_blocks(
  U_sf = U_sf,
  X_rast = X_rast,
  id_col = "U_id",
  verbose = FALSE
)

pred_U <- cos_predict_areal(
  fit = fit,
  rec_B = rec_B,
  U_blocks = U_blocks,
  method = "mean",
  verbose = TRUE
)

summary(pred_U)
```

Prediction functions default to `target = "latent"`. Use
`target = "observed"` when the prediction should include support-level nugget
variation on the requested support.

## Cross validation

`cos_cv_observed()` performs K-fold cross validation by refitting the model to
training plots and predicting held-out plot polygons. By default it scores the
observed response. In simulations with known latent truth, it can score both
latent and observed targets from the same refits.

```{r cv, eval = FALSE}
set.seed(1)
fold_id <- sample(rep(seq_len(10), length.out = nrow(B_sf)))

cv <- cos_cv_observed(
  fit = fit,
  X_rast = X_rast,
  B_sf = B_sf,
  target = c("latent", "observed"),
  latent_col = "eta_B_true",
  k = 10,
  fold_id = fold_id,
  fit_args = list(n_batch = 50, batch_length = 8),
  recover_args = list(burn_in = 100, n_samples = 60),
  seed = 1,
  verbose = FALSE
)

summary(cv)
```

For model-comparison guidance, see the spatial-guidance vignette. For model
and computational details, see the model-details vignette.

## Reference

Zhang, L., Finley, A. O., Nothdurft, A., and Banerjee, S. (2024). Bayesian
modeling of incompatible spatial data: A case study involving Post-Adrian storm
forest damage assessment. *International Journal of Applied Earth Observation
and Geoinformation*, 135, 104224. <https://doi.org/10.1016/j.jag.2024.104224>
