---
title: "Getting Started with cosTapered"
author: "Andrew O. Finley, Lu Zhang, Arne Nothdurft, and Sudipto Banerjee"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with cosTapered}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Overview

`cosTapered` fits a Bayesian change-of-support model for data where responses
are observed on polygons, covariates are available on a raster, and predictions
are needed for raster cells or other polygons. Typical forest inventory uses
include plot-level field observations, remotely sensed raster covariates, and
stand- or management-unit predictions.

The main workflow is:

1. Build the raster covariate stack exactly as the model should see it.
2. Prepare observed-support weights with `cos_prepare()`.
3. Specify priors with `cos_default_priors()`.
4. Fit covariance parameters with `cos_fit()`.
5. Recover observed-support latent quantities with `cos_recover_B()`.
6. Predict fine-support or areal-support quantities with
   `cos_predict_fine()` and `cos_predict_areal()`.

The package does not add an intercept, scale covariates, buffer points, choose
the taper distance `gamma`, or convert areal means to totals. Those choices are
left to the analyst.

## Example data

The package includes a small simulated example with a canopy-height raster,
observed plot polygons, areal prediction polygons, a forest boundary, and
known simulation truth.

```{r packages}
library(sf)
library(raster)
library(ggplot2)
library(cosTapered)
```

```{r read-data}
chm <- raster(system.file("extdata", "example_chm.tif", package = "cosTapered"))
names(chm) <- "chm"

B_sf <- st_read(system.file("extdata", "example_B.gpkg",
                            package = "cosTapered"), quiet = TRUE)
U_sf <- st_read(system.file("extdata", "example_U.gpkg",
                            package = "cosTapered"), quiet = TRUE)
forest_sf <- st_read(system.file("extdata", "example_forest.gpkg",
                                 package = "cosTapered"), quiet = TRUE)
truth <- readRDS(system.file("extdata", "example_truth.rds",
                             package = "cosTapered"))

truth[c("beta_true", "tau_B_sq_true", "sigma_sq_true",
        "eff_range_true", "phi_true", "gamma", "n_B", "n_U")]
```

Users prepare the raster covariates. Here we add an intercept raster and keep
the canopy-height layer unscaled.

```{r build-raster-design}
intercept <- chm
intercept[] <- ifelse(is.na(getValues(chm)), NA_real_, 1)
names(intercept) <- "intercept"

X_rast <- stack(intercept, chm)
names(X_rast) <- c("intercept", "chm")
```

```{r input-plot}
chm_df <- as.data.frame(chm, xy = TRUE, na.rm = TRUE)
names(chm_df) <- c("x", "y", "chm")

ggplot() +
  geom_tile(data = chm_df, aes(x = x, y = y, fill = chm),
            width = xres(chm), height = yres(chm)) +
  geom_sf(data = U_sf, fill = NA, color = "white", linewidth = 0.25) +
  geom_sf(data = forest_sf, fill = NA, color = "black", linewidth = 0.55) +
  geom_sf(data = B_sf, fill = NA, color = "#c1121f", linewidth = 0.45) +
  scale_fill_viridis_c(option = "cividis") +
  coord_sf(expand = FALSE) +
  labs(title = "Raster covariate, observed plots, and prediction polygons",
       fill = "CHM", x = NULL, y = NULL) +
  theme_bw()
```

## Prepare supports

`cos_prepare()` computes area weights linking observed polygons to raster
cells, observed-support covariates, the observed-support nugget matrix, and
the distance/taper information needed for spatial fitting.

This chunk is not evaluated during vignette build because it does geometry
overlay work, but it is the code to run in an analysis.

```{r prepare, eval = FALSE}
prep <- cos_prepare(
  X_rast = X_rast,
  B_sf = B_sf,
  response_col = "y_B",
  gamma = truth$gamma,
  taper_code = truth$taper_code,
  n_threads = 1,
  verbose = FALSE
)

summary(prep)
```

## Specify priors

Priors are specified on the covariate scale supplied by the user. The package
uses `tau_B_sq`, the average observed-support nugget variance, because this is
usually easier to think about than the fine-support nugget variance `tau_sq`.

```{r priors, eval = FALSE}
priors <- cos_default_priors(
  prep = prep,
  beta_mu = c(intercept = 40, chm = 1),
  beta_sd = c(intercept = 20, chm = 2),
  tau_B_shape = 2,
  tau_B_scale = truth$tau_B_sq_true,
  sigma_shape = 2,
  sigma_scale = truth$sigma_sq_true,
  phi_lower = 3 / 600,
  phi_upper = 3 / 80
)
```

The exponential effective range is approximately `3 / phi`. Thus
`phi_lower = 3 / 600` and `phi_upper = 3 / 80` allow effective ranges from
about 80 to 600 map units.

## Fit the model

`cos_fit()` samples the collapsed posterior for covariance parameters. Use
longer chains and multiple starting values for real analyses.

```{r fit, eval = FALSE}
fit <- cos_fit(
  prep = prep,
  priors = priors,
  n_chains = 3,
  starting = data.frame(
    tau_B_sq = truth$tau_B_sq_true * c(0.5, 1, 2),
    sigma_sq = truth$sigma_sq_true * c(0.5, 1, 2),
    phi = truth$phi_true * c(0.75, 1, 1.25)
  ),
  tuning = c(log_tau_B_sq = 0.25, log_sigma_sq = 0.25, z_phi = 0.25),
  n_batch = 100,
  batch_length = 25,
  seed = 1,
  verbose = TRUE
)

summary(fit, burn_in = 50)
plot(fit, burn_in = 50)
```

Trace plots should move across stable regions after burn-in. If `phi` piles up
against a prior bound, revisit the effective range bounds or the taper
distance `gamma`.

## Recover and predict

`cos_recover_B()` samples regression coefficients, observed-support spatial
effects, and observed-support latent means conditional on retained covariance
parameter draws.

```{r recover, eval = FALSE}
rec_B <- cos_recover_B(
  fit = fit,
  burn_in = 50,
  n_samples = 100,
  seed = 1,
  verbose = FALSE
)

summary(rec_B, include_eta_B = TRUE)
```

Fine-support prediction returns the latent raster-scale response by default.

```{r predict-fine, eval = FALSE}
pred_fine <- cos_predict_fine(
  fit = fit,
  rec_B = rec_B,
  method = "mean",
  verbose = FALSE
)

summary(pred_fine)
```

Areal prediction first builds prediction-support blocks and then predicts
support averages.

```{r predict-areal, eval = FALSE}
U_blocks <- cos_make_areal_blocks(
  U_sf = U_sf,
  X_rast = X_rast,
  id_col = "U_id",
  verbose = FALSE
)

pred_U <- cos_predict_areal(
  fit = fit,
  rec_B = rec_B,
  U_blocks = U_blocks,
  method = "mean",
  verbose = TRUE
)

summary(pred_U)
```

Prediction functions default to `target = "latent"`. Use
`target = "observed"` when the prediction should include a new independent
nugget draw on the requested support.

## Cross validation

`cos_cv_observed()` performs K-fold cross validation by refitting the model to
training plots and predicting held-out plot polygons. By default it scores the
observed response. In simulations with known latent truth, it can score both
latent and observed targets from the same refits.

```{r cv, eval = FALSE}
set.seed(1)
fold_id <- sample(rep(seq_len(10), length.out = nrow(B_sf)))

cv <- cos_cv_observed(
  fit = fit,
  X_rast = X_rast,
  B_sf = B_sf,
  target = c("latent", "observed"),
  latent_col = "eta_B_true",
  k = 10,
  fold_id = fold_id,
  fit_args = list(n_batch = 50, batch_length = 8),
  recover_args = list(burn_in = 100, n_samples = 60),
  seed = 1,
  verbose = FALSE
)

summary(cv)
```

For model-comparison guidance, see the spatial-guidance vignette. For model
and computational details, see the model-details vignette.
