---
title: "Getting Started with cosTapered"
subtitle: "Gaussian change-of-support fitting, fine-grid prediction, and areal summaries"
author: "Andrew O. Finley, Lu Zhang, Arne Nothdurft, and Sudipto Banerjee"
format:
  html:
    toc: true
    toc-location: left
    toc-expand: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    code-overflow: wrap
    html-math-method: katex
    css: cosTapered-vignette.css
execute:
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{Getting Started with cosTapered}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Overview

`cosTapered` fits Bayesian change-of-support (COS) models for spatial data
whose responses, covariates, and prediction targets are measured on different
supports. In the COS setting considered here, covariates are represented on a
fine raster grid, responses are observed on polygons, and predictions can be
made for either fine raster cells or other polygon supports.

The fitting and prediction chunks below are displayed without execution to
keep vignette builds quick. Run them in order for the complete workflow; the
short chains illustrate the API and require longer runs and convergence
diagnostics for inference.

Development of the package was motivated by data structures common in natural
resource and forest inventory applications, where field responses are measured
on plots, remotely sensed covariates are available as raster layers, and
management decisions often require predictions for stands or other irregular
areal units. The model follows the tapered COS formulation used by Zhang et
al. (2024) for incompatible spatial forest damage data.

The main workflow is:

1. Supply the raster covariate stack used to form the design matrix.
2. Prepare observed-support weights with `cos_prepare()`.
3. Specify priors with `cos_default_priors()`.
4. Fit covariance parameters with `cos_fit()`.
5. Recover observed-support latent quantities with `cos_recover_B()`.
6. Predict fine-support or areal-support quantities with
   `cos_predict_fine()` and `cos_predict_areal()`.

The workflow keeps data preparation choices explicit. The raster stack defines
the design matrix, including any intercept or transformed covariates; observed
supports are supplied as polygons; the taper distance `gamma` is chosen before
fitting; and areal predictions are returned as support averages, leaving any
conversion to totals to the analysis context.

## Example data

The package includes a small simulated example with a canopy-height raster,
observed plot polygons, areal prediction polygons, a forest boundary, and
known simulation truth.

```{r packages}
library(sf)
library(raster)
library(ggplot2)
library(cosTapered)
```

```{r read-data}
chm <- raster(system.file("extdata", "example_chm.tif", package = "cosTapered"))
names(chm) <- "chm"

B_sf <- st_read(system.file("extdata", "example_B.gpkg",
                            package = "cosTapered"), quiet = TRUE)
U_sf <- st_read(system.file("extdata", "example_U.gpkg",
                            package = "cosTapered"), quiet = TRUE)
forest_sf <- st_read(system.file("extdata", "example_forest.gpkg",
                                 package = "cosTapered"), quiet = TRUE)
truth <- readRDS(system.file("extdata", "example_truth.rds",
                             package = "cosTapered"))

truth[c("beta_true", "tau_B_sq_true", "sigma_sq_true",
        "eff_range_true", "phi_true", "gamma", "n_B", "n_U")]
```

## Supply the raster design

The model design matrix is taken from the layers in the raster stack. Here the
stack contains an intercept layer and the canopy-height layer.

```{r build-raster-design}
intercept <- chm
intercept[] <- ifelse(is.na(getValues(chm)), NA_real_, 1)
names(intercept) <- "intercept"

X_rast <- stack(intercept, chm)
names(X_rast) <- c("intercept", "chm")
```

```{r input-plot}
chm_df <- as.data.frame(chm, xy = TRUE, na.rm = TRUE)
names(chm_df) <- c("x", "y", "chm")

ggplot() +
  geom_tile(data = chm_df, aes(x = x, y = y, fill = chm),
            width = xres(chm), height = yres(chm)) +
  geom_sf(data = U_sf, fill = NA, color = "white", linewidth = 0.25) +
  geom_sf(data = forest_sf, fill = NA, color = "black", linewidth = 0.55) +
  geom_sf(data = B_sf, fill = NA, color = "#c1121f", linewidth = 0.45) +
  scale_fill_viridis_c(option = "cividis") +
  coord_sf(expand = FALSE) +
  labs(title = "Raster covariate, observed plots, and prediction polygons",
       fill = "CHM", x = NULL, y = NULL) +
  theme_bw()
```

The observed plot responses are measured on the plot polygons. The COS model
connects these plot-support responses to the raster covariates through
support averaging.

```{r response-plot}
ggplot() +
  geom_sf(data = forest_sf, fill = "grey95", color = "grey40", linewidth = 0.3) +
  geom_sf(data = U_sf, fill = NA, color = "white", linewidth = 0.25) +
  geom_sf(data = B_sf, aes(fill = y_B), color = "grey15", linewidth = 0.25) +
  scale_fill_viridis_c(option = "magma") +
  coord_sf(expand = FALSE) +
  labs(title = "Observed Gaussian plot-support responses",
       fill = "y_B", x = NULL, y = NULL) +
  theme_bw()
```

## Prepare observed supports

`cos_prepare()` computes area weights linking observed polygons to raster
cells, observed-support covariates, the observed-support nugget matrix, and
the distance/taper information needed for spatial fitting.

This is the first model-building step after the raster design and observed
polygons have been assembled.

By default, `cos_prepare()` stops if incomplete raster covariate cells would
leave an observed-support row sum materially below one. This protects the
support-average target. Use `missing = "drop"` or
`missing = "renormalize"` only when the intended analysis should use covered
cells rather than the full observed support.

```{r prepare, eval = FALSE}
prep <- cos_prepare(
  X_rast = X_rast,
  B_sf = B_sf,
  response_col = "y_B",
  gamma = truth$gamma,
  taper_code = truth$taper_code,
  n_threads = 1,
  verbose = FALSE
)

summary(prep)
```

## Specify priors

Priors are specified on the covariate scale supplied by the user. The package
uses `tau_B_sq`, the average observed-support nugget variance, because this is
usually easier to think about than the fine-support nugget variance `tau_sq`.

```{r priors, eval = FALSE}
priors <- cos_default_priors(
  prep = prep,
  beta_mu = c(intercept = 40, chm = 1),
  beta_sd = c(intercept = 20, chm = 2),
  tau_B_shape = 2,
  tau_B_scale = truth$tau_B_sq_true,
  sigma_shape = 2,
  sigma_scale = truth$sigma_sq_true,
  phi_lower = 3 / 600,
  phi_upper = 3 / 80
)
```

The exponential effective range is approximately `3 / phi`. Thus
`phi_lower = 3 / 600` and `phi_upper = 3 / 80` allow effective ranges from
about 80 to 600 map units.

## Fit the model

`cos_fit()` samples the collapsed posterior for covariance parameters. Use
longer chains and multiple starting values for real analyses. `burn_in` is
the first retained iteration in each chain, not a number of batches;
`burn_in = 50` below retains iteration 50 onward.

```{r fit, eval = FALSE}
fit <- cos_fit(
  prep = prep,
  priors = priors,
  n_chains = 3,
  starting = data.frame(
    tau_B_sq = truth$tau_B_sq_true * c(0.5, 1, 2),
    sigma_sq = truth$sigma_sq_true * c(0.5, 1, 2),
    phi = truth$phi_true * c(0.75, 1, 1.25)
  ),
  tuning = c(log_tau_B_sq = 0.25, log_sigma_sq = 0.25, z_phi = 0.25),
  n_batch = 100,
  batch_length = 25,
  seed = 1,
  verbose = TRUE
)

summary(fit, burn_in = 50)
plot(fit, burn_in = 50)
```

Trace plots should move across stable regions after burn-in. If `phi` piles up
against a prior bound, revisit the effective range bounds or the taper
distance `gamma`.

## Recover observed-support quantities

`cos_recover_B()` samples regression coefficients, observed-support spatial
effects, and observed-support latent means conditional on retained covariance
parameter draws.

```{r recover, eval = FALSE}
rec_B <- cos_recover_B(
  fit = fit,
  burn_in = 50,
  n_samples = 100,
  seed = 1,
  verbose = FALSE
)

summary(rec_B, include_eta_B = TRUE)
```

## Predict the fine grid

Fine-support prediction returns the latent raster-scale response by default.
If `pred_coords` and `X_pred` are omitted, `cos_predict_fine()` predicts only
on the complete raster cells intersecting the observed polygons, stored by
`cos_prepare()`. To map the full raster domain, explicitly supply every
complete cell as below. The example uses the native 3 m by 3 m grid.

For routine pixel-wise maps with uncertainty, use
`spatial_uncertainty = "marginal"`. This uses the faster marginal prediction
backend: each fine cell receives its own one-dimensional conditional Gaussian
draw for each retained posterior sample, without forming a dense joint
covariance matrix across all cells. The resulting cell means, posterior
standard deviations, and intervals are appropriate when cells are interpreted
one at a time. Use `spatial_uncertainty = "joint"` only when a downstream
summary combines cells and requires cross-cell covariance.

```{r predict-fine, eval = FALSE}
pred_cell <- which(complete.cases(getValues(X_rast)))
pred_coords <- xyFromCell(X_rast[[1]], pred_cell)
X_pred <- as.matrix(extract(X_rast, pred_cell))
colnames(X_pred) <- names(X_rast)

pred_fine <- cos_predict_fine(
  fit = fit,
  rec_B = rec_B,
  pred_coords = pred_coords,
  X_pred = X_pred,
  spatial_uncertainty = "marginal",
  keep_samples = FALSE,
  n_threads = 2,
  verbose = FALSE
)

summary(pred_fine)
```

`pred_fine$summary` contains one row per predicted fine cell, including the
cell coordinates, posterior mean, and posterior standard deviation.
`summary(pred_fine)` summarizes those results across cells. Set
`keep_samples = TRUE` to retain draws and add the 2.5%, 50%, and 97.5%
posterior quantiles; interval summaries are not returned with
`keep_samples = FALSE`. The same rule applies to areal predictions.

## Predict areal supports

Areal prediction first builds prediction-support blocks and then predicts
support averages. `cos_make_areal_blocks()` uses the same missing-covariate
policy as `cos_prepare()`. For per-stand or per-polygon estimates with valid
one-unit-at-a-time uncertainty, use `spatial_uncertainty = "marginal"`.

```{r predict-areal, eval = FALSE}
U_blocks <- cos_make_areal_blocks(
  U_sf = U_sf,
  X_rast = X_rast,
  id_col = "U_id",
  verbose = FALSE
)

pred_U <- cos_predict_areal(
  fit = fit,
  rec_B = rec_B,
  U_blocks = U_blocks,
  spatial_uncertainty = "marginal",
  keep_samples = FALSE,
  n_threads = 2,
  verbose = TRUE
)

summary(pred_U)
```

Prediction functions default to `target = "latent"`. Use
`target = "observed"` when the prediction should include support-level nugget
variation on the requested support. For areal prediction,
`spatial_uncertainty = "marginal"` returns marginal samples for each
prediction unit, not joint samples across units; do not use those samples for
totals, contrasts, or rankings that require cross-unit covariance.

## Cross validation

`cos_cv_observed()` performs K-fold cross validation by refitting the model to
training plots and predicting held-out plot polygons. By default it scores the
observed response. In simulations with known latent truth, it can score both
latent and observed targets from the same refits.

```{r cv, eval = FALSE}
set.seed(1)
fold_id <- sample(rep(seq_len(10), length.out = nrow(B_sf)))

cv <- cos_cv_observed(
  fit = fit,
  X_rast = X_rast,
  B_sf = B_sf,
  target = c("latent", "observed"),
  latent_col = "eta_B_true",
  k = 10,
  fold_id = fold_id,
  fit_args = list(n_batch = 50, batch_length = 8),
  recover_args = list(burn_in = 100, n_samples = 60),
  seed = 1,
  verbose = FALSE
)

summary(cv)
```

For model-comparison guidance, see the [spatial-guidance
vignette](cosTapered-spatial-guidance.qmd). For model and computational
details, see the [model-details vignette](cosTapered-model.qmd).

## Reference

Zhang, L., Finley, A. O., Nothdurft, A., and Banerjee, S. (2024). Bayesian
modeling of incompatible spatial data: A case study involving Post-Adrian storm
forest damage assessment. *International Journal of Applied Earth Observation
and Geoinformation*, 135, 104224. <https://doi.org/10.1016/j.jag.2024.104224>
